---
title: Correlation Matrix
author: Brian Helsel
date: '2022-07-27'
slug: correlation-matrix
categories: []
tags: []
---

## Introduction

Correlations are one of the simplest ways to look at the association between 2 variables. However, I often find myself running correlations on many variables when exploring a new data set. I was tired of writing `cor(x, y)` and I wanted a quick way to visualize correlations, so I created a correlation matrix using `ggplot2`.

## Data

Let's explore the mtcars data set and create a correlation matrix.

```{r load data and libraries, message=FALSE}

library(dplyr)
library(magrittr)
library(ggplot2)
library(reactable)

data("mtcars")

reactable(
  mtcars, 
  defaultPageSize = 5,
  columns = list(
    .rownames = colDef(
      name = "model", 
      sortable = TRUE, 
      minWidth = 200)
    )
  )

```
<br>
<center>
<hr>
In the mtcars data set above, there are **`r ncol(mtcars)` columns** and **`r nrow(mtcars)` rows** of data. 

The columns include: **`r names(mtcars)`**. 

Here is a description of all the variables.

| Variable | Description |
| :------: | :----------: |
| mpg | Mile per gallon |
| cyl | Number of Cylinders |
| disp | Displacement |
| hp | Gross Horsepower |
| drat | Rear Axle Ratio |
| wt | Weight (lb/1000) |
| qsec | Quarter Mile Time |
| vs | Straight Engine |
| am | Manual Transmission |
| gear | Number of Forward Gears |
| carb | Number of Carburetors | 
</center>

## The Final Product

This is the correlation matrix that we will be creating and I've added the code to a function at the end of this post. For those interested in learning, I've provided a description of the key steps in getting the correlations and processing the data before creating the correlation matrix. Otherwise, feel free to jump ahead to the bottom of the post to see a link to the `cor_matrix` function. 

```{r, echo=FALSE, fig.width=8, fig.height=8}


vars <- c("mpg", "cyl", "disp", "hp", "drat", "wt", 
          "qsec", "vs", "am", "gear", "carb")

var.names <- c("Miles/Gallon", "Number of Cylinders", 
               "Displacement", "Gross Horsepower",
               "Rear Axle Ratio", "Weight (lb/1000)", 
               "Quarter Mile Time", "Straight Engine",
               "Manual Transmission", 
               "Number of Forward Gears", 
               "Number of Carburetors")

myplot <- bhelselR::cor_matrix(dataset = mtcars, vars = vars, var.names = var.names)
myplot[[1]]

```

## Correlations

To create the correlation matrix, we will use the `rcorr` function from the `Hmisc` package. First, we need to take the mtcars data set and convert it to a matrix before calling the `Hmisc::rcorr` function. We will use spearman correlations for this example, but the type argument can also be switched to pearson.

```{r running correlations}

vars <- c("mpg", "cyl", "disp", "hp", "drat", 
          "wt", "qsec", "vs", "am", "gear", "carb")

corres <- 
  mtcars %>%
  select(all_of(vars)) %>%
  as.matrix() %>%
  Hmisc::rcorr(., type = "spearman")

```

When we run the code above, you can see that `Hmisc::rcorr` returns a list. From this list, we can extract the correlation coefficients and their corresponding p-values. You can extract this list to a matrix using the `extract2` function from the `magrittr` package or an alternative method if you prefer. We can also round each value to 2 digits to make the output cleaner.

```{r extract rho and pvalues}

cor.matrix.r <- 
  corres %>%
  extract2("r") %>%
  round(., 2)

cor.matrix.r

cor.matrix.p <- 
  corres %>%
  extract2("P") %>%
  round(., 2)

cor.matrix.p

```

## Data Processing

Now you have the correlations and the p-values for all the variables. However, as you can see from the correlation matrix, you have repeat values for each variable with diagonal 1.000 values for correlations between the same variables. When viewing the correlation matrix, it is not helpful to see these repeat values so we can set them to missing and exclude them from the visualization using the `lower.tri` function from `base` R.

```{r set repeat values to missing}
cor.matrix.r[lower.tri(cor.matrix.r)] <- NA
cor.matrix.r
cor.matrix.p[lower.tri(cor.matrix.p)] <- NA
cor.matrix.p
```

It is almost time to visualize the data. We have 2 final steps to complete the data processing. First, let's rename the variables to something more meaningful. This way when others see them on the plot, it will be easier to understand. We will need to rename the columns with `colnames` and rows with `rownames`. We can also reshape the data using `melt` from the `reshape2` package. Adding the `na.rm = TRUE` argument removes all the duplicate correlation values that were set to missing earlier. 

```{r rename variables and reshape data}

var.names <- c("Miles/Gallon", 
               "Number of Cylinders", 
               "Displacement", 
               "Gross Horsepower",
               "Rear Axle Ratio", 
               "Weight (lb/1000)", 
               "Quarter Mile Time", 
               "Straight Engine",
               "Manual Transmission", 
               "Number of Forward Gears", 
               "Number of Carburetors")

colnames(cor.matrix.r) <- c(var.names)
rownames(cor.matrix.r) <- c(var.names)
colnames(cor.matrix.p) <- c(var.names)
rownames(cor.matrix.p) <- c(var.names)

cor.matrix.r <- 
  cor.matrix.r %>% 
  reshape2::melt(., na.rm = TRUE)

head(cor.matrix.r, 10)
  
```

## The Plot

Now all we need to do is to write the code to visualize the data. Here I will use `geom_tile` which creates a box for each value and fills it with color based on the value variable from the data set. We can use `scale_fill_gradient2` function to set the low, mid, and high colors for the correlation plot. The low and high values would correspond to -1 and 1, respectively. We can also rotate the x-axis labels by 45 degrees to make them easier to read and add the correlation values to the plot using `geom_text`. All of the other lines of code are to adjust the theme of the plot and can be changed to your preferences.

```{r plot part 1, fig.height=8, fig.width=8}

ggplot(cor.matrix.r, aes(x = Var2, y = Var1, fill=value)) +
  geom_tile(color="black") +
  scale_fill_gradient2(low="#B2BDED", mid="#ffffff", high="#C6E0B4",
                       midpoint=0, limit=c(-1,1), space="Lab", 
                       name="Spearman\nCorrleation") +
  theme_minimal() +
  coord_fixed() +
  geom_text(aes(Var2, Var1, label = value, fontface = "bold"), 
            color = "black", size = 4) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1, 
                                   face = "bold"),
        axis.text.y = element_text(size = 12, face = "bold"),
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid.major = element_blank(),
        panel.border = element_blank(), 
        panel.background = element_blank(), 
        axis.ticks = element_blank(),
        legend.justification = c(1, 0), 
        legend.position = c(0.5, 0.7), 
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5))

```

You can also add the p-values to a Table. Here I only display p-values < 0.05 if `Var1` is *Miles/Gallon*, but you could display all p-values or adjust the filters as needed.

```{r pvalue tables}

cor.matrix.p %>%
  reshape2::melt(., na.rm = TRUE) %>%
  filter(value < 0.05 & Var1 == "Miles/Gallon") %>%
  mutate(value = ifelse(value < 0.001, "< 0.001", value)) %>%
  gt::gt() %>%
  gtExtras::gt_theme_538()

```

## The Function

It can be all combined into a single function. You can view the function in the <a href="https://github.com/bhelsel/bhelselR/blob/master/R/visualizations.R">visualization.R</a> file my personal GitHub page or download it into your RStudio by typing the following code into your console.

`devtools::install_github("bhelsel/bhelselR")` 

That's it! Fairly simple and straightforward. There are also a lot of different ways that you could modify this function to personalize it for your analyses. Let me know if you use this function or a similar one in the comments below!